---
title: "Assignment 2 - Language Development in ASD - Making predictions"
author: "Lina Pedersen, Thea Pedersen, Sophie Korch, Andreas Jeppesen"
date: "August 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the second exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time against some new data. Does the model generalize well?
Then we will learn to do better by cross-validating models and systematically compare them.

The questions to be answered (in a separate document) are:
1- Discuss the differences in performance of your model in training and testing data
2- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
3- Predict a new kid's performance (let's call him Bernie) and discuss it against expected performance of the two groups

## Learning objectives

- Critically appraise the predictive framework (contrasted to the explanatory framework)
- Learn the basics of machine learning workflows: training/testing, cross-validation, feature selections

## Let's go

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/5pc05mh5jwvdfjk/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/eegu8fea2entdqv/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/cf4p84mzn2p1bev/token_test.csv?dl=0

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())


```{r, include = FALSE}

pacman::p_load(readr,dplyr,stringr,lmerTest,Metrics,caret, tidyverse)

#Loading data
Demo <- read_csv("demo_test.csv")
LU <- read_csv("LU_test.csv")
Word <- read_csv("token_test.csv")

## Clean up function, included to inspire you

# I have changed some names in the function to make it fit our cleaned up training data

CleanUpData <- function(Demo, LU, Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      SUBJ = SUBJ, 
      VISIT=VISIT) %>%
    mutate(
      VISIT = as.numeric(str_extract(VISIT, "\\d")),
      SUBJ = gsub("\\.","", SUBJ)
      ) %>%
    dplyr::select(
      SUBJ, VISIT, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    rename(
      SUBJ = Child.ID,
      VISIT = Visit) %>% 
    dplyr::select(
      SUBJ, VISIT, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      SUBJ = gsub("\\.","", SUBJ)
    )
    
  Data=merge(Demo,Speech,all=T)
  
  Data1= Data %>% 
     subset(VISIT=="1") %>% 
     dplyr::select(SUBJ, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(SeverityOfSymptoms = ADOS, 
            VerbalIQ1 = ExpressiveLangRaw, 
            NonVerbalIQ1 = MullenRaw,
            SocialSkills1 = Socialization) 
  
  Data=merge(Data, Data1, all=T) %>%
    mutate(
      SUBJ = as.numeric(as.factor(as.character(SUBJ))),
      VISIT = as.numeric(as.character(VISIT)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "ASD",
         "B"  = "TD")
    )

  return(Data)
}

# We now use the function on our three test datasets and save it as a combined test dataset
testdf <- CleanUpData(Demo, LU, Word)

# 
write_csv(testdf, "testdf.csv")

# Load training Data
traindf <- read_csv("autismdata.csv")

#- recreate the models you chose last time (just write the code again and apply it to Train Data)
m1 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + VerbalIQ1, data = traindf, REML = FALSE)


#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())

# We get the predicted values from the model:
fitteddf <- data.frame(fitted(m1))

CHI_MLU <- traindf$CHI_MLU %>%  # saving the real CHI_MLU vlaues
  na.exclude()             # in which we exclude the NA's


fitteddf$CHI_MLU <- CHI_MLU   # Adding a column to the predicted.df with the CHI_MLU values

# Finding the rmse value
rmse(fitteddf$CHI_MLU , fitteddf$fitted.m1.)

#- test the performance of the models on the test data (Tips: google the functions "predict()")

# First we exclude NAs from the test data:
testdf <- subset(testdf, is.na(testdf$CHI_MLU) == FALSE)

p1 <- data.frame(predict(m1, testdf, allow.new.levels = TRUE)) %>% 
  rename( predictedCHI_MLU = predict.m1..testdf..allow.new.levels...TRUE.)
p1


# We test the performance of the model on the test data with the rmse measure:
rmse(testdf$CHI_MLU , p1$predictedCHI_MLU)

#- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

```

[HERE GOES YOUR ANSWER]

### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model (Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)


- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data


```{r}
#- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).

# We make a subset of the training data with only relevant variables in which we omit NAs:
subtraindf <- select(traindf, "SUBJ", "VISIT", "Diagnosis", "CHI_MLU", "VerbalIQ1", "SeverityOfSymptoms1") %>% 
  na.omit()

# I create the model
mbasic <- lmerTest::lmer(CHI_MLU ~ VISIT + Diagnosis + (1 + VISIT | SUBJ),  data = subtraindf, REML = FALSE)

#- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)

# We will make a 6-fold crossvalidation, so k is 6
k = 6

# Creating the folds:
folds = createFolds(unique(subtraindf$SUBJ, k = k, list = TRUE, returnTrain = FALSE))

trainRMSE = rep(NA, k)
testRMSE = rep(NA, k)

i = 1
for (fold in folds) {
  train = subset(subtraindf, !(SUBJ %in% fold))
  test = subset(subtraindf, SUBJ %in% fold)
  model = lmerTest::lmer(CHI_MLU ~ VISIT + Diagnosis + (1 + VISIT | SUBJ),  data = train, REML = FALSE)
  test$prediction = predict(model, test, allow.new.levels = TRUE)
  train$prediction = fitted(model)
  trainRMSE[i] = rmse(train$CHI_MLU, fitted(model))
  testRMSE[i] = rmse(test$CHI_MLU, test$prediction)
  i = i + 1
}

# Getting RMSE values for train and test folds:
trainRMSE
testRMSE
# Getting means of the values:
mean(trainRMSE)
mean(testRMSE)

#- Report the results and comment on them.
### The average of the numbers from trainRMSE is 0.34 which is considerably lower than the average of the values of testRMSE (0.76). This means that the model is not very good at predicting new data.

#- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

# Creating new folds:
folds = createFolds(unique(subtraindf$SUBJ, k = k, list = TRUE, returnTrain = FALSE))

# Resetting i to 1 and running the loop again with the m1 model that we found best in assignment 2 part 1:
i = 1
for (fold in folds) {
  train = subset(subtraindf, !(SUBJ %in% fold))
  test = subset(subtraindf, SUBJ %in% fold)
  model = lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + VerbalIQ1 + (1 + VISIT | SUBJ),  data = train, REML = FALSE)
  test$prediction = predict(model, test, allow.new.levels = TRUE)
  train$prediction = fitted(model)
  trainRMSE[i] = rmse(train$CHI_MLU, fitted(model))
  testRMSE[i] = rmse(test$CHI_MLU, test$prediction)
  i = i + 1
}

# Getting RMSE values for train and test folds:
trainRMSE
testRMSE
# Getting means of the values:
mean(trainRMSE)
mean(testRMSE)

# The average of the test RMSE values is now 0.56 which is much closer to the mean of the train RMSE (0.33). Therefore, the model we found last week is better than the basic model.

# Creating new folds:
folds = createFolds(unique(subtraindf$SUBJ, k = k, list = TRUE, returnTrain = FALSE))

# Resetting i to 1 and running the loop again with a model like before but Severity of Symptoms added as an extra predictor:
i = 1
for (fold in folds) {
  train = subset(subtraindf, !(SUBJ %in% fold))
  test = subset(subtraindf, SUBJ %in% fold)
  model = lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + VerbalIQ1 + SeverityOfSymptoms1 + (1 + VISIT | SUBJ),  data = train, REML = FALSE)
  test$prediction = predict(model, test, allow.new.levels = TRUE)
  train$prediction = fitted(model)
  trainRMSE[i] = rmse(train$CHI_MLU, fitted(model))
  testRMSE[i] = rmse(test$CHI_MLU, test$prediction)
  i = i + 1
}

# Getting RMSE values for train and test folds:
trainRMSE
testRMSE
# Getting means of the values:
mean(trainRMSE)
mean(testRMSE)

# The mean test RMSE is now 0.55, so we conclude that it does not make enough of a difference to include SeverityOfSymptoms as a predictor as we do not want to overfit either.

# Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
# Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data
```

[HERE GOES YOUR ANSWER]

### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
# Subsetting Bernies data:
berniedf <- subset(testdf, SUBJ == 5)

# How does he compare to TD children at each visit?
# Finding the mean for each visit for TD children only:
MeanMLUTD <- subtraindf %>% 
  filter(Diagnosis == "TD") %>% 
  group_by(VISIT) %>% 
  summarise(
    meanTD = mean(CHI_MLU)
  )


# Adding the list of means to Bernies df for comparison:
berniedf$CHI_MLUmeans <- MeanMLUTD$meanTD

# Difference between means and Bernie's values:
berniedf$CHI_MLUdifference <- berniedf$CHI_MLU - berniedf$CHI_MLUmeans

# We take a look at the absolute differences:
abs(berniedf$CHI_MLUdifference)


# We make a plot to illustrate Bernie's values compared to the mean values:
berniedf %>% 
  ggplot() +
  aes(x = berniedf$CHI_MLUmeans, y = berniedf$CHI_MLU) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "Red") +
  theme_minimal() +
  xlab("TD Means of Child MLU") +
  ylab("Bernie's values of MLU")


# How does Bernie do compared to the prediction of the model?
# We predict Bernies data:

p2 <- predict(m1, berniedf)

# We take a look at the model's predicted values for Bernie:
p2

# The predicted value for visit 6 is 3.6.
# Bernie's value for Visit 6 is 2.4, thus Bernie is way below the predicted value.

```


[HERE GOES YOUR ANSWER]

### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

